Stochastic Blockmodels for Relational Event Data

### 2012-05-29 docs/submission.tex
Fix up inference section.
Rearrange results/discussion a bit.

\subsection*{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
% \begin{align*}
% p(z_r | \mathbf{z}_{-r},\mathcal{A}_t,\alpha,\boldsymbol{\beta}) &\propto p(z_i | \mathbf{z}_{-i},\alpha) p(\mathcal{A}_t|\mathbf{z},\boldsymbol{\beta}) \\
% p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta},\mbox{node} \ r \ \mbox{involved}) &= \prod_{m=1}^M \lambda_{i_a,j_a}(t_a|\cdot)
% \prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
% \end{align*}
\begin{align*}
p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
%& \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t - t_{M}) \lambda_{ij}(t|\cdot)\}  
\end{align*} 
where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.  This aims for similarly sized groups.}
% \begin{align*}
% p(z_i = k | z_{-i},\alpha) &= n_k & p(z_i = \mbox{new} |  z_{-i},\alpha) &= \alpha
% \end{align*}

@@

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\ %, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) \\  
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
%p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}, \mbox{cluster} \ k \ \mbox{or} \ l \ \mbox{involved}) &= \prod_{a \in \mathcal{U}_r} \lambda_{i_a,j_a}(t_a|\cdot)
%\prod_{(i,j) \in \mathcal{R}_{i_a,j_a}} \exp \{ -(t_a - \tau_{aij}) \lambda_{ij}(t_a|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.  This distribution is sampled using slice sampling.

\subsection*{Sampling $\mu$  and $\sigma$ given $\beta$ }

The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
\begin{align}
\label{eqn:gibbs.sigma}
\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
 \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
\end{align}
The posterior distribution of $\mu$ given $\sigma$ and the lower level parameters can in turn be written as
\begin{align}
\label{eqn:gibbs.mu}
\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K}\sum_{k=1}^K \beta_{k,p},\frac{ \sigma_p^2}{\sqrt{K}}\right),
\end{align}

%$\boldsymbol{\phi}_{l,r}^{split} \sim G_0$, $\boldsymbol{\phi}_{r,l}^{split} \sim G_0$, and $\boldsymbol{\phi}_{l,l}^{split} \sim G_0$ (or actually $N(\mu_{\phi},\Sigma_{\phi})$

% TODO: Update the following discussion about sampling.

% This operation can be done in parallel, and only the portion of the likelihood involving the newly assigned cluster  need to be recomputed.  Computing the likelihood is then $O(T \cdot P \cdot 4N)$ (though this computation is split into $K^2$ pieces) and must computed $N \cdot K$ times per Gibbs iteration.  The most straightforward implementation has space complexity $O(N^2)$ since we must keep track of $\tau_{i,j,m}$ for each $(i,j)$ at a given $m$.

% \subsection{Approximate likelihood methods}

% TODO: Discuss how one might sample the likelihood.  Cite Raftery paper.

% TODO: Look at performance of this with respect to bias and performance measure

@@

\section{Model checking and experiments}

@@

@@

 Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.  For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.  Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.

For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.  This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).  \footnote{Will need to discuss results for other datasets once they are complete.}

Figure \ref{fig:recall} will provide recall curves for several of the datasets.

@@

 We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.  The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models) to provide detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in dynamics of the network as a whole.

Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann data. Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.

@@

This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share.  Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.

>>>>

\subsection{Sampling $\mathbf{z}$ given $\boldsymbol{\beta}$ and $\mathcal{A}$ }

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]}  \prod_{(i,j) \in \mathcal{R}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\} 
\end{align*} 
where $\mathcal{R}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.  Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = \mbox{new} |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.  In our experiments we use $\alpha=0.1$ and use Algorithm 8 from \cite{Neal2001} with 5 extra clusters drawn from the prior.\footnote{I also tried out a prior for cluster sizes such that $n_k \sim \mbox{NegBinom}(3,N/K)$, where $K$ is chosen a priori.  This aims for similarly sized groups.}

@@

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\ 
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} 
\prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.  This distribution is sampled using slice sampling.

\subsection{Sampling $\mu$  and $\sigma$ given $\beta$ }

The inverse gamma distribution is a conjugate prior to a Gaussian distribution with known location parameter, thus $\sigma$ can be sampled from the conditional distribution
\begin{align}
\label{eqn:gibbs.sigma}
\sigma_p^2 &| \boldsymbol{\beta}, \mu_p,  \alpha_{\sigma}, \beta_{\sigma} \sim
 \mbox{Inv-Gamma}\left(\alpha_{\sigma} + \frac{K^2}{2}, \beta_{\sigma} + \frac{1}{2} \sum_{k=1}^K\sum_{l=1}^K (\beta_{p,k,l} - \mu_p)^2\right).
\end{align}
Each parameter $\mu_p$ can be sampled from its conditional distribution given $\sigma_p^2$ and the parameters $\boldsymbol{\beta}$, 
\begin{align}
\label{eqn:gibbs.mu}
\mu_p | \boldsymbol{\beta},\sigma_p^2 &\sim \mbox{Normal}\left(\frac{1}{K^2}\sum_{k=1}^K \sum_{l=1}^K \beta_{p,k,l},\frac{ \sigma_p^2}{\sqrt{K^2}}\right).
\end{align}

@@

\section{Model fitting and experiments}

@@

\subsection{Exploratory analysis}

Figure \ref{fig:posteriorparams} shows examples of model estimates from data sets of dyadic interactions for the Eckmann email data. On the left we

@@

TODO: Quickly describe main take-aways from the results tables.

% Table \ref{tab:results} compares the train and test likelihood for each method and data set combination.  For the synthetic data generated with $K=2$ clusters, the model with $K=2$ is best as expected.  Fitting the model with a single cluster also outperforms the simple baselines with respect to both likelihoods.

%For the Eckmann subset the fitted model with $K=2$ performs best except for the test data using the conditional logit likelihood.  This may be because the model has overfit the training set, or the sampler has not properly converged (as is the case with $K=3$).  \footnote{Will need to discuss results for other datasets once they are complete.}

@@

 We propose a hierarchical approach for modeling event-based network data that is analogous to recent hierarchical extensions for latent position models \cite{Handcock2007} and exponential random graph models \cite{Schweinberger2011}.  The method combines a latent variable framework (stochastic blockmodels) with a local dependence model for event sequences (relational event models \cite{Butts2008}).  This provides detailed models of event dynamics among subsets of nodes while allowing for heterogeneity in dynamics of the network as a whole.

%Panels contain posterior credible intervals for parameters pertaining to dyads belonging to a particular block.

@@

This analysis suggests key differences exist in typical behaviors or roles that subsets of nodes share across a variety of data sets.  Other theories could be explored by including relevant statistics in the specification of $\mathbf{s}(t,i,j,\mathcal{A}_t)$, and similarly one can use our method to study how the roles of these statistics vary across nodes.
